Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2018-07-03 writeup/response2reviews.tex

\reply{
As we discussed in reponse to the other reviewer,
we intentionally chose simulations in this in-between spot;
they are somewhat biologically realistic 
while still retaining a clear way of obtaining the ``right'' answer 
(with regard to admixture proportions and number of discrete populations).
Additional sets of simulations will, in our opinion, make the paper too lengthy and complex;
the applications to real data we hope demonstrate the real-world utility of the method.
We have done ``sanity check'' simulations under the model itself,
but as these are far enough from biological reality (e.g., the ``frequencies'' are not between 0 and 1),
we think they'd only make the paper more confusing.
As for the migration rates -- these are given in the text already \llname{ll:sim_params},
and we have added an explicit reference to this \llname{ll:sim_params_ref}.
}

@@

\begin{point}{}
    There seem to be some possible degeneracy issues, 
for instance when $\alpha_d$ goes to 0 (which seems to be the case for some of the inferred layers), 
$\alpha_0$ and $\phi_k$ will becomes indistinguishable.
 You should mention these issues, and if you have dealt with them 
 (or think they are not important, as you are not really interested in estimating these parameters).
\end{point}

\reply{We agree: in practice,
because those parameters can't be easily interpreted in a biological context, 
we treat them more as nuisance parameters and focus on inference of the admixture proportions.
We have plans for a future release that has a more biologically motivated model of IBD (see more below).
We've made a note about this. \revref}

\begin{point}{}
    Genetics states ``All articles published by GENETICS must include a 
Data Availability Statement at the end of the Materials and Methods section." 
You don't have such a section at the moment.
For reproducibility of your results, 
you should make at least the covariance matrix and the positions of each deme you used in your analysis available. 
Ideally share all of the genetic data and the scripts used to filter it, including the settings (this is often quite important). 
But if someone is protective of this data, I personally would be fine with omitting it. But this is the editors call to make.
\end{point}

\reply{Thanks for pointing this out.  
We have added a Data Availability Statement, 
and all data analyzed and scripts used to perform the analyses will be archived in Data Dryad.
\revref
}

### 2018-07-03 writeup/conStruct.tex
minor edits

\section*{Data availability}
The method \texttt{conStruct} is implemented as an R package, 
and is available for installation at \burl{https://github.com/gbradburd/conStruct}.
Scripts for generating and analyzing all simulated and empirical datasets, 
as well as the datasets themselves, 
are also available at the same site. \revpoint{2}{11}

@@

To test the method, we first generated data using the coalescent simulator \texttt{ms} \cite{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past, 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}),
and collapsed the layers together using random, spatially autocorrelated admixture proportions.
See \secref{sim_details} for more details, including parameter values used. \linelabel{ll:sim_params_ref}
We simulated datasets using $K=1$, 2, and 3 layers;
in each simulation we sampled 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation with 10 replicates.
For comparison,
we also analyzed each simulated dataset using ADMIXTURE \cite{ADMIXTURE} with $K$ between 1 and 7, 
and compared models using ADMIXTURE's cross-validation procedure with 50 folds.

@@

For a relatively small number of samples, 
\texttt{conStruct} can be much faster than existing nonspatial Bayesian clustering methods.
On a desktop machine, using a single 4.2 GHz Intel Core i7 processor, 
an analysis of the black bear dataset (78 samples, 21,000 loci) 
running \texttt{conStruct}'s spatial model with 4 layers for 5,000 MCMC iterations 
(which was more than sufficient for convergence) 
took 2.8 hours.
On the same machine and dataset, 
a fastSTRUCTURE \cite{fastStructure} run with $K=4$ took 90.2 hours.
However, for a large number of samples, 
fastSTRUCTURE might be faster.
For almost any size dataset,
the maximum likelihood algorithm implemented in ADMIXTURE 
is quite a bit faster than \texttt{conStruct}:
running ADMIXTURE on the bear dataset over all values of $K$ from 1 to 7, 
including 50-fold cross-validation for each value of $K$, took only 6.6 minutes.
It should also be noted that there may be situations when the
binomial-based model underlying ADMXITURE performs better than
our Gaussian-based model,
e.g., when clusters differ at only a few, strongly differentiated loci
\linelabel{ll:admixture_few_loci}.

@@

\paragraph{Implications for management and conservation}
Because isolation by distance is common, 
a likely result of applying \texttt{conStruct} to existing data is that
populations previously identified as distinct using nonspatial clustering methods 
may be grouped into the same layer.
This ``lumping'' might better reflect the demographic history of these populations,
but may not contradict the genetic distinctness implied by the nonspatial clustering.
This genetic distinctness rather than shared history may be more relevant for
management decisions and conservation policy,
which are often predicated on the identification of discrete ``management units'' (MUs) 
identified using genetic data \cite{Moritz1994,Waples_1998,Moritz_etal_2002}.

@@

A second caveat is that, in some instances, 
membership in the same layer may not mean that samples are particularly related.
If covariance within a layer decays sharply with distance, 
and the layer-specific relatedness parameter $\phi^{(k)}$ is low, 
individuals separated by a large spatial distance may be in the same layer but have very low pairwise relatedness.
It is possible that this is happening in Fig \ref{populus_sp_pies}.
At $K=3$, the southernmost populations of \textit{P. trichocarpa} are in the gold layer, 
whose other neighbors are to the north, with an intervening group of populations in the red layer, 
and at $K=5$, those southernmost samples split out and become their own layer.
Furthermore, note that in this case $\alpha_0^{(k)}$ and $\phi^{(k)}$ are confounded,
so differences in $\phi$ between layers should not be overinterpreted. \revpoint{2}{10}
Again, we encourage users to run analyses across multiple values of $K$, 
to examine the spatial covariance functions within layers when interpreting results.

>>>>

\section*{Data availability}
The method \texttt{conStruct} is implemented as an R package, 
and is available for installation at \burl{https://github.com/gbradburd/conStruct}.
Scripts for generating and analyzing all simulated and empirical datasets, 
as well as the datasets themselves, 
are also available at the same site, 
and additionally have been archived at Data Dryad (doi:XXXXX)  \revpoint{2}{11}

@@

To test the method, we first generated data using the coalescent simulator \texttt{ms} \cite{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past, 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}),
with randomly distributed but spatially autocorrelated admixture proportions.
See \secref{sim_details} for more details, including parameter values used. \linelabel{ll:sim_params_ref}
We simulated datasets using $K=1$, 2, and 3 layers;
in each simulation we sampled 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation with 10 replicates.
For comparison,
we also analyzed each simulated dataset using ADMIXTURE \cite{ADMIXTURE} with $K$ between 1 and 7, 
and compared models using ADMIXTURE's cross-validation procedure with 50 folds.

@@

For a relatively small number of samples, 
\texttt{conStruct} can be much faster than existing nonspatial Bayesian clustering methods.
On a desktop machine, using a single 4.2 GHz Intel Core i7 processor, 
an analysis of the black bear dataset (78 samples, 21,000 loci) 
running \texttt{conStruct}'s spatial model with 4 layers for 5,000 MCMC iterations 
(which was more than sufficient for convergence) 
took 2.8 hours.
On the same machine and dataset, 
a fastSTRUCTURE \cite{fastStructure} run with $K=4$ took 90.2 hours.
However, for a large number of samples, 
fastSTRUCTURE might be faster.
For almost any size dataset,
the maximum likelihood algorithm implemented in ADMIXTURE 
is quite a bit faster than \texttt{conStruct}:
running ADMIXTURE on the bear dataset over all values of $K$ from 1 to 7, 
including 50-fold cross-validation for each value of $K$, took only 6.6 minutes.
It should also be noted that there may be situations when the
binomial-based model underlying ADMXITURE performs better than
our Gaussian-based model,
e.g., when clusters differ at only a few strongly differentiated loci
\linelabel{ll:admixture_few_loci}.

@@

\paragraph{Implications for management and conservation}
Because isolation by distance is common, 
a likely result of applying \texttt{conStruct} to existing data is that
populations previously identified as distinct using nonspatial clustering methods 
may be grouped into the same layer.
This ``lumping'' might better reflect the demographic history of these populations,
but may not contradict the genetic distinctness implied by the nonspatial clustering.
This genetic distinctness -- rather than shared history -- may be more relevant for
management decisions and conservation policy,
which are often predicated on the identification of discrete ``management units'' (MUs) 
identified using genetic data \cite{Moritz1994,Waples_1998,Moritz_etal_2002}.

@@

A second caveat is that, in some instances, 
membership in the same layer may not mean that samples are particularly related.
If covariance within a layer decays sharply with distance, 
and the layer-specific relatedness parameter $\phi^{(k)}$ is low, 
individuals separated by a large spatial distance may be in the same layer but have very low pairwise relatedness.
It is possible that this is happening in Fig \ref{populus_sp_pies}.
At $K=3$, the southernmost populations of \textit{P. trichocarpa} are in the gold layer, 
whose other neighbors are to the north, with an intervening group of populations in the red layer, 
and at $K=5$, those southernmost samples split out and become their own layer.
Furthermore, note that in this case $\alpha_0^{(k)}$ and $\phi^{(k)}$ are confounded,
so differences in $\phi$ between layers should not be overinterpreted. \revpoint{2}{10}
Again, we encourage users to run analyses across multiple values of $K$ 
and to examine the spatial covariance functions within layers when interpreting results.
